---
title: "Methods and Results"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library("tidyverse")
library("knitr")
library("tidytext")
library("lubridate")
library("stargazer")
library("wordcloud")
library("kableExtra")
library("plotly")
```
<br>

```{r, include=FALSE}
dat <- read_csv("data/processed_data.csv")
sw_regex <- paste0(" ", paste0(stop_words$word ,collapse = " | ")) %>% paste0(" s")

bing <- get_sentiments("bing")
```

```{r}
dem_articles <- dat %>% 
  filter(president == "dem")

republican_articles <- dat %>% 
  filter(president == "rep")
```

## Headline Word Count Distribution

```{r, echo=FALSE}
fig <- dem_articles %>% 
  plot_ly(x = ~headline_word_count, type="histogram", visible=T, alpha=0.6, nbinsx=20, color = I("blue"))
fig <- fig %>% add_histogram(x = ~republican_articles$headline_word_count, type = "histogram", visible = F, color = I("red"))

fig <- fig %>% layout(
  title = "Headline Word Count Distribution During Different Time Periods",
  xaxis = list(title = "Headline Word Count"),
  yaxis = list(title = "No. Articles"),
  updatemenus = list(
    list(
    y = 0.9,
    buttons = list(
      list(method = "restyle",
           args = list("visible", list(TRUE, FALSE)),
           label = "2013-2016"),
      list(method = "restyle",
           args = list("visible", list(FALSE, TRUE)),
           label = "2017-2020")
    )
  ))
) %>% hide_legend()

fig
```

The above plot show the word count in NYT headlines distributions during Obama's and Trump's presidencies. 

## NYT Headlines Word Clouds {.tabset}

### 2013-2016 Word Cloud

```{r, echo=FALSE}
top50_common_dem <- dem_articles[, 1, drop = FALSE] %>% 
  unnest_tokens(word, title) %>% 
  anti_join(stop_words, by = c("word" = "word")) %>% 
  filter(word != "0091") %>% 
  count(word, sort = TRUE) %>% 
  top_n(50, n)
wordcloud(top50_common_dem$word, top50_common_dem$n)
```

### 2017-2020 Word Cloud

```{r, echo=FALSE}
top50_common_rep <- republican_articles[, 1, drop = FALSE] %>% 
  unnest_tokens(word, title) %>% 
  anti_join(stop_words, by = c("word" = "word")) %>% 
  filter(word != "0091") %>% 
  count(word, sort = TRUE) %>% 
  top_n(50, n)
wordcloud(top50_common_rep$word, top50_common_rep$n)
```

## {-}

The above word clouds show the 50 most frequently used words in NYT headlines during Obama's presidency (2013 - 2016) and during Trump's presidency (2017 - 2020). The size of the words in the word clouds reflects how often those words were used.

```{r}
top5_sections <- dat %>% 
  group_by(section_name) %>% 
  summarize(n = n()) %>% 
  arrange(-n) %>% 
  top_n(5, n)

top5_sections <- top5_sections$section_name

word_counts <- dat[, c(1, 2), drop = FALSE] %>% 
  filter(section_name %in% top5_sections) %>% 
  unnest_tokens(token, title) %>% 
  anti_join(stop_words, by = c("token" = "word")) %>% 
  count(token, section_name, sort=TRUE) %>% 
  filter(token != "0091")
```

## Most Common Headline Words in Top 5 NYT Categories {.tabset}

### U.S. Section

```{r}
word_counts %>% 
  dplyr::filter(section_name == top5_sections[1]) %>% 
  top_n(10, n) %>% 
  ggplot(mapping = aes(x = fct_reorder(token, n), y = n)) +
  geom_col(fill = "orange") +
  labs(x = "Word", y = "Number of Occurences in the Headlines") +
  coord_flip()
```

### Opinion Section

```{r}
word_counts %>% 
  dplyr::filter(section_name == top5_sections[2]) %>% 
  top_n(10, n) %>% 
  ggplot(mapping = aes(x = fct_reorder(token, n), y = n)) +
  geom_col(fill = "orange") +
  labs(x = "Word", y = "Number of Occurences in the Headlines") +
  coord_flip()
```


### World Section

```{r}
word_counts %>% 
  dplyr::filter(section_name == top5_sections[3]) %>% 
  top_n(10, n) %>% 
  ggplot(mapping = aes(x = fct_reorder(token, n), y = n)) +
  geom_col(fill = "orange") +
  labs(x = "Word", y = "Number of Occurences in the Headlines") +
  coord_flip()
```


### Arts Section

```{r}
word_counts %>% 
  dplyr::filter(section_name == top5_sections[4]) %>% 
  top_n(10, n) %>% 
  ggplot(mapping = aes(x = fct_reorder(token, n), y = n)) +
  geom_col(fill = "orange") +
  labs(x = "Word", y = "Number of Occurences in the Headlines") +
  coord_flip()
```

### Business Day Section
```{r}
word_counts %>% 
  dplyr::filter(section_name == top5_sections[5]) %>% 
  top_n(10, n) %>% 
  ggplot(mapping = aes(x = fct_reorder(token, n), y = n)) +
  geom_col(fill = "orange") +
  labs(x = "Word", y = "Number of Occurences in the Headlines") +
  coord_flip()
```

## {-}

The above plots show the frequency of the most commonly used words in NYT headlines for each of the 5 most common sections in the newspapers. These plots reflect only the headlines written between 2013 and 2020.

## Emotional Word Usage {.tabset}

### Obama's Presidency

```{r}
p <- dat %>% 
  filter(section_name %in% top5_sections) %>%
  filter(president == "dem") %>%
  pivot_longer(cols = c(negative_prop, positive_prop)) %>% 
  rename(Proportion = value) %>% 
  mutate(name = ifelse(name == "negative_prop", "Negative", "Positive"),
         Proportion = signif(Proportion, 3)) %>% 
  ggplot(mapping = aes(x = name, y = Proportion, fill = name)) +
  geom_bar(stat = "summary", fun = "mean") +
  facet_wrap(~section_name) +
  theme(legend.position = "none") +
  labs(x = "Sentiment", y = "Average Proportion in Headline")

ggplotly(p, tooltip = c("Proportion"))
```

### Trump's Presidency

```{r}
p <- dat %>% 
  filter(section_name %in% top5_sections) %>%
  filter(president == "rep") %>%
  pivot_longer(cols = c(negative_prop, positive_prop)) %>% 
  rename(Proportion = value) %>% 
  mutate(name = ifelse(name == "negative_prop", "Negative", "Positive"),
         Proportion = signif(Proportion, 3)) %>% 
  ggplot(mapping = aes(x = name, y = Proportion, fill = name)) +
  geom_bar(stat = "summary", fun = "mean") +
  facet_wrap(~section_name) +
  theme(legend.position = "none") +
  labs(x = "Sentiment", y = "Average Proportion in Headline")

ggplotly(p, tooltip = c("Proportion"))
```

## {-}

The above two plots show the percentage that negative and positive words used up in headlines during Obama's and Trump's presidencies. These percentages are shown separately for each section.

## Emotional Language Difference Confidence Intervals {.tabset}

```{r}
negative_conf_inters <- matrix(nrow = 0, ncol = 3, dimnames = list(c(), c("section", "lower", "upper")))
positive_conf_inters <- matrix(nrow = 0, ncol = 3, dimnames = list(c(), c("section", "lower", "upper")))
count <- 1

for (section in top5_sections) {
  current_section <- dat %>% 
    filter(section_name == section)
  
  curr_neg_inters <- t.test(current_section %>% filter(president == "dem") %>% select(negative_prop), current_section %>% filter(president == "rep") %>% select(negative_prop))$conf.int
  
  negative_conf_inters <- rbind(negative_conf_inters, c(section, curr_neg_inters[1], curr_neg_inters[2]))

  curr_positive_inters <- t.test(current_section %>% filter(president == "dem") %>% select(positive_prop), current_section %>% filter(president == "rep") %>% select(positive_prop))$conf.int
  
  positive_conf_inters <- rbind(positive_conf_inters, c(section, curr_positive_inters[1], curr_positive_inters[2]))
}

negative_conf_inters <- negative_conf_inters %>% 
  as.data.frame()
negative_conf_inters$lower <- as.double(negative_conf_inters$lower)
negative_conf_inters$upper <- as.double(negative_conf_inters$upper)
negative_conf_inters <- negative_conf_inters %>%
  mutate(point_estimate = (lower + upper) / 2)

positive_conf_inters <- positive_conf_inters %>% 
  as.data.frame()
positive_conf_inters$lower <- as.double(positive_conf_inters$lower)
positive_conf_inters$upper <- as.double(positive_conf_inters$upper)
positive_conf_inters <- positive_conf_inters %>%
  mutate(point_estimate = (lower + upper) / 2)

negative_conf_inters <- negative_conf_inters %>% 
  mutate(sd = upper - point_estimate)

positive_conf_inters <- positive_conf_inters %>% 
  mutate(sd = upper - point_estimate)
```

### Negative Word Usage

```{r}
negative_conf_inters %>% 
  plot_ly(x = ~point_estimate, y = ~section, type = "scatter", mode = "markers",
          error_x = ~list(array = sd),
          hoverinfo = "text",
          text = ~paste(
            "Section Name:", section, "\n",
            "Point Estimate:", signif(point_estimate, 3), "\n",
            "CI Lower Bound:", signif(lower, 3), "\n",
            "CI Upper Bound:", signif(upper, 3), "\n"
          )) %>% 
  layout(xaxis = list(title = "Proportional Difference (Dem - Rep)"),
         yaxis = list(title = "Section Name"))
```

### Positive Word Usage

```{r}
positive_conf_inters %>% 
  plot_ly(x = ~point_estimate, y = ~section, type = "scatter", mode = "markers",
          error_x = ~list(array = sd),
          hoverinfo = "text",
          text = ~paste(
            "Section Name:", section, "\n",
            "Point Estimate:", signif(point_estimate, 3), "\n",
            "CI Lower Bound:", signif(lower, 3), "\n",
            "CI Upper Bound:", signif(upper, 3), "\n"
          )) %>% 
  layout(xaxis = list(title = "Proportional Difference (Dem - Rep)"),
         yaxis = list(title = "Section Name"))
```

## {-}

The plots above show the confidence intervals for the difference in negative and positive word usages between Obama's presidency and Trump's presidency. 